---
title: "EDRM878 Assignment5"
author: Tiejun Zhang
output: html_notebook
---

### 9.7. After the explosion of the space shuttle Challenger on January 28, 1986, a study was conducted [1, 4] to determine if previously-collected data about the ambient air temperature at the time of launch could have been used to foresee potential problems with the launch (Table 4.1; data set: shuttles).In Example 4.2, a model was proposed for these data.

```{r}
data(shuttles)
head(shuttles)
summary(shuttles)
```


#### 1. Plot the data.
```{r}
plot(shuttles$Damaged/6 ~ shuttles$Temp,
     las = 1,
     main = "Proportion of Damaged O-rings and Temperatures", 
     ylab = "Proportion of damaged O-rings",
     xlab = "Temperature")
```
The O-rings tend to fail more often at low temperatures.

#### 2. Fit and interpret the proposed model.
```{r}
# fit the model
shuttle.model <- glm(Damaged/6 ~ Temp,
                data = shuttles,
                family = binomial,
                weights=rep(6, length(Temp)))

# the effect of temperature on O-ring failure
coef(shuttle.model)

# the temperature drops by 1 Fahrenheit
exp(-coef(shuttle.model)[2])

# the temperature drops by 10 Fahrenheit
exp(-10*coef(shuttle.model)[2])

```
The odds of O-ring failure increase by 1.12times for every 1 Fahrenheit drop in temperature, and the odds increase by 3.18 times if the temperature drops 10 Fahrenheit.

#### 3. Perform a diagnostic analysis.

Check for Overdispersion
```{r}
c(Df = df.residual(shuttle.model),
  Resid.dev = deviance(shuttle.model),
  Pearson.X2 = sum(resid(shuttle.model, type = "pearson")^2))
```
The residual deviance is smaller than the residual degrees of freedom and Pearson statistics is a bit greater than the residual degrees of freedom, not enough evidence to support overdispersion.


Quantile Residuals vs Fitted Values
```{r}
qres <- qresid(shuttle.model)
scatter.smooth(qres ~ sqrt(fitted(shuttle.model)),
               ylab = "Quantile residuals",
               xlab = "Sqrt(Fitted values)",
               las = 1)
```
Q-Q plot
```{r}
qqnorm(qres, las = 1)
abline(0, 1)
```
The Q-Q plot looks acceptable.



Outliers
```{r}

rs <- cbind(rD = resid(shuttle.model),
            "r'D" = rstandard(shuttle.model),
            "r''" = rstudent(shuttle.model),
            rQ = qresid(shuttle.model))

head(rs)

apply(abs(rs), 2, max)
```
The maximum all four types of resduals are greater than 2.5, but this shoud not be a big concern as suggeted by the Q-Q plot.



Influential observations
```{r}
im <- influence.measures(shuttle.model)
colSums(im$is.inf)
rowSums(im$is.inf)
plot(cooks.distance(shuttle.model), type = "h", main = "Cook's distance", ylab = "D", xlab = "Observation number", las = 1)

```
1 observations have high leverage, as identified by the column labelled hat, 4 observations are identified by the covariance ratio as influential, 1 observation is identified by the DIFFITS as influential, but Cook's distance does not identify any observation as influential.


#### 4. On the day of the Challenger launch, the forecast temperature was 31???F. What is the predicted probability of an O-ring failure?
```{r}
predict(model.logit, newdata=data.frame(Temp=31), type="response")
```


#### 5. What would the ed50 mean in this context? What would be a more sensible ed for this context?

```{r}
library(MASS)

ED50s <- dose.p(shuttle.model)

ED50s
```
The ED50 refers to the temperature(44 Fahrenheit) at which 50% of the O-rings fail. In this case we want the O-ring failures as less as possible, a lower ED, which means less O-ring failures, is perfered.


### 10.12. In a study of depressed women [15], women were classified into groups (Table 10.15; data set: dwomen) based on their depression level (Depression), whether a severe life event had occurred in the last year (SLE), and if they had three children under 14 at home (Children). Model these counts using a Poisson glm, and summarize the data if possible.

#### Data Preparation
```{r}
data(dwomen)
head(dwomen)
summary(dwomen)
```
#### Removing structural zeros
```{r}
dwomen1 <- subset(dwomen, !(Depression == "Yes" & SLE == "No" & Children == "Yes"))
```


#### Frequncy Tables
```{r}

dwomen1.tab <- xtabs(Counts ~ Depression + SLE + Children,
               data = dwomen1)
dwomen1.tab
```


#### Variable Relations
Let us take a look at the relationship between the response and the explanatory variables.
```{r}
plot(Counts ~ Depression, data = dwomen1)
plot(Counts ~ SLE, data = dwomen1)
plot(Counts ~ Children, data = dwomen1)

```
It seems that $Depression$ and $Children$ have effects on $Counts$. Observations with depression or children tend to have more counts. The effect of $SLE$ on $Counts$ is not obvious.


#### Fit Model
Let us start with the full model that includes all interactions and see what happens if we drop the three-way interaction.
```{r}
gm.1 <- glm(Counts ~ Depression * SLE * Children,
            data = dwomen1,
            family = poisson)
drop1(gm.1, test="Chisq")
```
The three-way interaction is not necessary in the model. Let us consider dropping each twoway
interaction in turn:

```{r}
gm.2 <- update( gm.1, . ~ (Depression  + SLE+ Children)^2 )
drop1(gm.2, test="Chisq")
```
The interaction between $SLE$ and $Children$ is not necessary in the model.

Now we fit the model
```{r}
gm.3 <- glm(Counts ~ Depression * SLE + Depression * Children,
            data = dwomen1,
            family = poisson)

# coefficents and effect size
coef(gm.3)
exp(coef(gm.3))

```
The count of those who have depression and children is 11.5 times what it is for those who don't have depression and children.
The count of those who have depression and SLE is 4.1 times what it is for those who don't have depression and SLE.

#### Diagnostic Analysis
Check for Overdispersion
```{r}
c(Df = df.residual(gm.3),
  Resid.dev = deviance(gm.3),
  Pearson.X2 = sum(resid(gm.3, type = "pearson")^2))
```


Quantile Residuals vs Fitted Values
```{r}
#library(statmod)

qr <- qresid(gm.3)

scatter.smooth(qr ~ sqrt(fitted(gm.3)),
               ylab = "Quantile residuals",
               xlab = "Sqrt(Fitted values)",
               las = 1)
```
Q-Q plot
```{r}
qqnorm(qr, las = 1)
abline(0, 1)
```

Influential observations
```{r}
im <- influence.measures(gm.3)
colSums(im$is.inf)
rowSums(im$is.inf)
plot(cooks.distance(gm.3), type = "h", main = "Cook's distance", ylab = "D", xlab = "Observation number", las = 1)

```