---
title: "Final Project"
author: Tiejun Zhang
output: html_notebook
---

***

### Normal

The Normal distribution is used to model continuous data that have a symmetric distribution.

The mean-variance relationship: $V (\mu) = 1$

Dispersion: The dispersion $??$ is unknow and needs to be estimated. There are four types 
of estimators: MLE estimator, Modified profile log-likelihood estimator, Mean deviance 
estimator, Pearson estimator. MLE estimator is biased, unless $p'/n$ is very small, the 
modified profile estimator is difficult to compute. In practice, the Pearson estimator 
tends to be more variable (less precise) but less biased than the mean deviance estimator. 
For normal GLMs, the latter three estimators make little difference.The mean deviance and 
Pearson estimators are very convenient.The mean deviance estimator should behave well when 
the saddlepoint approximation holds; that is, for normal or inverse Gaussian glms or when
$??$ is relatively small.

Link function: $\eta = \mu$

Inference: Walds test, Likelihood ratio and score test with selected estimator.All 
tests are based on large-sample asymptotic results, which apply when n is reasonably large. 
Walds test is the most commonly used. Confidence intervals for $\hat{\mu}$ are found by 
first computing confidence intervals for $\hat{\eta}$, and then applying the inverse link 
function (that is, ?? = g???1(??)) to the lower and upper limit of the interval found for 
$\hat{\eta}$. Likelihood ratio deals with nested models. The score test evaluates the 
usefulness of predictors.

The saddlepoint approximation is exact for the normal distribution.

Diagnostic Analysis:
The three main types of residuals are raw residuals, standardized residuals, and Studentized 
residuals. The standardized and Studentized residuals have approximately constant variance 
of one, and are preferred in residual plots.
Check for independence of the responses when possible.
Check for linearity between the responses and all covariates 
Residuals against the fitted values
Q-Q plot
Cook's distance plot
Influential observations

Solutions
If the responses are not independent, use other methods.
If the variance of the response is not approximately constant, transform the response as 
necessary.
if the relationship is not linear, transform the covariates using simple transformations 
(Sect. 3.10), polynomials in the covariates (Sect. 3.11), or regression splines (Sect. 3.12).



#### Data discription
The data set is available as the *crawl* data in the GLMsData library.

The data come from a study which hypothesized that babies would take longer to 
learn to crawl in colder months because the extra clothing restricts their 
movement. From 1988 to 1991, researchers recorded the babies' first crawling age
and the average monthly temperature 6 months after birth (when ``infants 
presumably enter the window of locomotor readiness''). The parents reported the 
birth month and age when their baby first crept or crawled a distance of four 
feet in one minute. Data were collected at the University of Denver Infant Study
Center on 208 boys and 206 girls, and summarized by the birth month.


#### Variables

BirthMonth: the baby's birth month

Age: the mean age (in completed weeks) at which the babies born this month
started to crawl

SD: the standard deviation (in completed weeks) of the crawling ages for babies
born this month

SampleSize: the number of babies in the study born in the given month

Temp: the monthly average temperature (in degrees F) six months after the birth
month


#### Research question

What is the effect of temperature (six months after birth) on the age when babies start to 
crawl?

For every 1 (degree F) increase in temperature, the mean age for babies starting to crawl 
woiuld be 0.08 weeks (about half day) earlier.



#### Analysis

1. Prepare the data and take a look at the data structure
```{r}
library(GLMsData)
data(crawl)
str(crawl)
head(crawl)
```

2. Explore the relationship between the response and predictor.
```{r}
scatter.smooth(crawl$Temp,
               crawl$Age,las = 1,
               xlab = "Temperature (in degree F)",
               ylab = "Mean Age (in Weeks)")
```
It seems that the temperature has an effect on the mean age at which the babies born this 
month started to crawl. The mean age decreases as temperature drops down.


3. Let us model age with temperature.
```{r}
crawl.model <- glm(Age ~ Temp,
                  data = crawl,
                  weights = SampleSize,
                  family = gaussian)

printCoefmat(coef(summary(crawl.model)))
```
The model test results indicate that temperature indeed exert an infulence on the mean age 
at which the babies started to crawl. For every 1 (degree F) increase in temperature, the 
mean age woiuld decrease by 0.08 weeks. 

4. Dianostic Analysis

The residuals against the fitted values, the standardized residuals was used here.
```{r}
scatter.smooth(rstandard(crawl.model) ~ fitted(crawl.model))
```
Q-Q Plot
```{r}
qqnorm( rstandard( crawl.model ), las=1 )
qqline( rstandard( crawl.model ) )
```

Cook's distance
```{r}
plot( cooks.distance(crawl.model), type="h", las=1)
```
The plots look acceptable, but there is a lack of linearity.


Influential observations
```{r}
im <- influence.measures(crawl.model)
colSums(im$is.inf)
```
The influence diagnostics reveal that two observations are influential according
to covariance ratio, but none are influential according to Cook's distance.


#### Additional exercise

We learned that the Wald test, the likelihood ratio test, and the score test
all provide the same result for the normal EDM. Verify that this is the case
for these data and your model.

Wald test
```{r}
phi.meandev <- deviance(crawl.model) / df.residual(crawl.model)
phi.pearson <- summary(crawl.model)$dispersion

# Both the mean deviance and Pearson estimators are similar. 
# The mean deviance estimator is used.
printCoefmat(coef(summary(crawl.model, dispersion = phi.pearson)))
```

Likelihood ratio test
```{r}
anova(crawl.model, test = "F", dispersion = phi.meandev)
```

Score test
```{r}
library(statmod)
m0 <- glm(Age ~ 1,
                data = crawl,
                weights = SampleSize,
                family = gaussian)

t.Ht <- glm.scoretest(m0, crawl$Temp)
p.Ht <- 2 * pt(-abs(t.Ht), df = df.residual(m0))
tab <- data.frame(Score.stat = t.Ht, P.value = p.Ht)
print(tab, digits = 3)
```

All tests suggest that the predictor is effective (p < 0.05), but the score test is much 
more conservative (p = 0.041) than the Wald and Likelihood tests (p = 0.002).

***



### Binomial
The binomial GLM is used to model proportions.

The mean-variance relationship: $V (\mu) = \mu(1-\mu)$

Dispersion: The dispersion $??$ is known to be $1/m$, no estimation is necessary.However, 
there may be overdispersion issue.

Link function: 1. logit link function (canonical and default); 2. probit link function; 
3. complementary log-log link function.The logit and probit link functions are very similar, 
and both are symmetric about $\mu$ = 0.5, whereas the complementary log-log link function is not.

The ed50 is the value of the covariates when the expected proportion is $\mu$ = 0.5.

Inference: Wald test doesn't work well with binomial GLMs. The score or likelihood ratio 
tests must be used.

The saddlepoint aprroximation is adequate for binomial distribution when my ??? 3 and 
m(1 ??? y) ??? 3.

Overdispersion: compare residual deviance and the Pearson goodness-of-fit statistic to the
residual degrees of freedom. Substantial differences suggest overdispersion. Overdispersion 
may arises from: 1.the $m_i$ cases, of which observation $y_i$ is a proportion, are not 
independent. (solution: use hierachical models); 2.the Bernoulli cases, that $m_i$ make up 
observation, are positively correlated. (solution: Quasi-binomial models)

No Goodness-of-Fit for Binary Responses

Diagnostic analysis of the model:
quantile residuals against the fitted values
Q-Q plot
Cook's distance
influential observations.




#### Data

The data set is available as the *belection* data in the GLMsData library.

The data give the number of male and females candidates in the British general
election held April 9, 1992.


#### Variables

Region: the region in Britain

Party: the political party

Females: the number of female candidates

Males: the number of male candidates


#### Research question

Assuming that the current political parties and views of British voters are 
similar now as they were in 1992 (which may not be a reasonable assumption),
what factors can best be used to predict female participation as a candidate
in British elections?

The factor Party is an effective predictor. Compared to the conservative party, the odds 
of a female being nominated as a candidate in the Green Party is 3.16 greater. The odds in 
the Labour Party and Other Party is 2.5 greater and in the Liberty Demoncratic Party is 
2.78 grater than the consevative party.




#### Analysis

1.Prepare the data
```{r}
data(belection)
str(belection)
summary(belection)
```

Let us create a variable Total by summing all male and female candidates.
```{r}
belection$Total <- belection$Males + belection$Females
```

2. Explore the relationship between the response and predictors.

Here is a visual display to help us determine if region make a difference on the proportion 
of female candidates.

```{r}

plot(Females / Total ~ Region,
     data = belection,
     las = 1,
     ylim = c(0,1))

```
Overall, there are much more male candiates than female candiates in every region. Several 
regions have more than 20% female candidates while most regions are below 20%. 


Here is a visual display to help us determine if Party makes a difference.

```{r}

plot(Females / Total ~ Party,
     data = belection,
     las = 1,
     ylim = c(0,1))

```
The Conservative Party on average has 10%, the lowest proportion, female candidates and the 
Green Party has the highest 30% female candidates. The other three parties have 20% female 
candiates.


3. Fit model.

Let us start with the interaction model
```{r}
female.model <- glm(Females / Total ~ Party * Region, 
                    data = belection, 
                    family = binomial,
                    weights = Total)
anova(female.model, test = "Chisq")

```
The likelihood test suggests that only Party makes a difference in the proportion of female 
candidates.



```{r}
female.model2 <- glm(Females / Total ~ Party, 
                    data = belection, 
                    family = binomial,
                    weights = Total)
printCoefmat(coef(summary(female.model2)))
```
Party membership does make a difference!

Let us see how R dummy-coded the levels of Party
```{r}
contrasts(belection$Party)
```


Now let's look at the effect sizes.
```{r}
exp(coef(female.model2))
```
Conservative party was set as the reference level by R, so the odds of a female being 
nominated as a candidate in the Green Party is 3.16 greater than in the conservative Party. 
The odds in the Labour Party and Other Party is 2.5 greater and in the Liberty Demoncratic 
Party is 2.78 grater than in the consevative party. 



4.Diagnostic Analysis

Overdispersion
```{r}
c(Df = df.residual(female.model2),
Resid.Dev = deviance(female.model2),
Pearson.X2 = sum( resid(female.model2, type="pearson")^2 ))
```
There is not too much difference between the three statistics.


Quantile residual vs fitted values.

```{r}
qres <- qresid(female.model2)
scatter.smooth(qres ~ fitted(female.model2),
               las = 1,
               main = "Residuals vs fitted",
               xlab = "Fitted value",
               ylab = "Quantile residual")


```


Q-Q Plot
```{r}
qqnorm(qres, las=1 )
qqline(qres)
```

Cook's distance
```{r}
plot( cooks.distance(female.model2), type="h", las=1)
```
The plots are pretty good.


Influential observations
```{r}
im <- influence.measures(female.model2)
colSums(im$is.inf)
```
The influence diagnostics reveal that two observations are influential according
to DFFITS, but none are influential according to Cook's distance.


#### Additional questions

1. How do you check for overdispersion when modeling proportions?

To look for overdispersion we compare residual deviance and the Pearson goodness-of-fit 
statistic to the residual degrees of freedom. Substantial differences suggest overdispersion.

2. Is there evidence of overdispersion in your chosen model?

*No evidence of overdispersion. There are not much difference between the residual deviance 
or the Pearson goodness-of-fit statistic and the residual degrees of freedom.*

```{r}
c(Df = df.residual(female.model2),
  Resid.dev = deviance(female.model2),
  Pearson.X2 = sum(resid(female.model2, type = "pearson")^2))
```


3. Whether or not there is evidence of overdispersion in your model, how would you change 
your analysis if you **did** detect overdispersion?

Overdispersion can arise from two major causes. The probabilitiesf $\mu_i$ are not 
constant between observations. Alternatively the $m_i$ cases, of which observation $y_i$ 
is a proportion, are not independent., we can address this type of overdispersion by a 
hierachical model. If overdispersion arises when  $m_i$ Bernoulli cases, that make up 
observations $y_i$, are positively correlated, we could use quasi-binomial models.

***

### Poisson
The Poisson distribution is used to model count data.
Poisson glms can be used to model rates (such as counts of cancer cases per unit of 
population) by using a suitable offset in the linear predictor.

The mean-variance relationship: $V (\mu) = \mu)$

Dispersion: The dispersion $??$ is known to be 1, no estimation is necessary.However, there 
may be overdispersion issue.

Link function: logit link function (canonical and default).

log-linear model: all explanatory variables are factors, the data can be summarized as a 
contingency table.
Poisson regression model: If any of the explanatory variables are quantitative.


Structure zeros: Contingency tables may contain cells with zero counts, cells containing 
structural zeros must be removed from the analysis.

Inference: Wald test, score or likelihood ratio tests can be used.

The saddlepoint aprroximation is adequate for Poisson distribution when y ??? 3.

Overdispersion: occurs when the variation in the responses is greater than expected under 
the Poisson model. Possible causes are that the model is misspecified (in which case the 
model should be amended), the means are not constant, or the responses are not independent. 
(solutions: Negative binomial GLMS and Qusi-poisson models)


Diagnostic analysis:
Overdispersion
quantile residuals against the fitted values
Q-Q plot
Cook's Distance plot
Influential observation



The data set is available as the *Sociology Survey Data* in an RData file provided via Blackboard.

Graduate students in sociology and other disciplines were randomly selected from the 
graduate student population at one university. They were asked, "What is the most important 
way to reduce crime?"

#### Variables

Response: (1) increase penalties, (2) increase police force, (3) increase
social services, (4) none of these

Major: (1) sociology, (2) other

Gender: (1) male, (2) female

#### Research questions

How do sociology graduate students differ from other graduate students in their perceptions 
about dealing with crime? Do these perceptions differ based on gender?

The graudate students from other majors have much lower odds to select $social service$ 
and $none$ than those who from sociology. The odds ratio for selecting social service is 
0.06 and the odds ratio for none is 0.09. Gender does not affect the perceptions very much.

#### Analysis

1. Prepare the data.
```{r}
load("Sociology Survey Data.RData")
str(soc_survey)
head(soc_survey)
summary(soc_survey)
```


Recode variables as factors.
```{r}
soc_survey$response <- factor(soc_survey$response, 
                              levels = c(1, 2, 3, 4), 
                              labels = c("penalty", "police force", "social services", "none"))
soc_survey$major <- factor(soc_survey$major, 
                           labels = c("Sociology", "Other"))
soc_survey$gender <- factor(soc_survey$gender, 
                            labels = c("male", "female"))
```

Let us calculate counts for each condition cell.
```{r}

ss.counts <- table(soc_survey)

ss.counts <- data.frame(ss.counts)

names(ss.counts) <- c("Response", "Major", "Gender", "Counts") 
ss.counts
```

Remove the cell that contains structure zero
```{r}
ss.counts1 <- subset(ss.counts, ss.counts$Counts != 0)
ss.counts1
```


Here it is as a contingency table.

```{r}
gm.table <- xtabs(Counts ~ Response+ Major  + Gender, data = ss.counts1)
gm.table
```

2. Take a look at the relationship between response and predictors.
```{r}
with(ss.counts1,{
  plot( Counts ~ Response, ylab="Counts", las=1)
  plot(Counts ~ Major, ylab="Counts", las=1)
  plot( Counts ~ Gender, ylab="Counts", las=1)
})
```



3. Fit the model.

Let us start with the saturated model.
```{r}
ss.m1 <- glm(Counts ~  Major * Gender * Response,
             data = ss.counts1,
             family = poisson)

anova(ss.m1, test = "Chisq")


```
The likelihood test suggests that Major has an interaction with response, but gender does 
not make a difference.


```{r}
ss.m2 <- glm(Counts ~ Major * Response,
             data = ss.counts1,
             family = poisson)

anova(ss.m2, test = "Chisq")
```

The left over (residual) deviance is 5.799 and the associated df is 7. Let's see if this 
is substantially different than what we expect for df = 7.

```{r}

pchisq(5.799, 7, lower.tail = FALSE)

```
Overdispersion model is not required.


Here are the coefficients.

```{r}
printCoefmat(coef(summary(ss.m2)))
```
The effect sizes.
```{r}
exp(coef(ss.m2))
```

The graudate students from other majors have much lower odds to select $social service$ 
and $none$ than those who from sociology. The odds ratio for selecting social service is 
0.06 and the odds ratio for none is 0.09.


4.Diagnostic Analysis

Overdispersion
```{r}
c(Df = df.residual(ss.m2),
Resid.Dev = deviance(ss.m2),
Pearson.X2 = sum( resid(ss.m2, type="pearson")^2 ))
```
There is not too much difference between the three statistics.


Quantile residual vs fitted values.

```{r}
qres <- qresid(ss.m2)
scatter.smooth(qres ~ fitted(ss.m2),
               las = 1,
               main = "Residuals vs fitted",
               xlab = "Fitted value",
               ylab = "Quantile residual")


```


Q-Q Plot
```{r}
qqnorm(qres, las=1 )
qqline(qres)
```

Cook's distance
```{r}
plot( cooks.distance(ss.m2), type="h", las=1)
```
The plots are not bad.


Influential observations
```{r}
im <- influence.measures(ss.m2)
colSums(im$is.inf)
```
The influence diagnostics reveal none observations are influential.


#### Additional questions

Is there any evidence of Simpson's paradox in these data? How do you know?

No, because there is only one interaction in the model and it doesn't depend on any other factors.
***

### Gamma

The Gamma distribution is used to model positive continuous data.Positive continuous data often have the variance increasing with increasing mean; the inverse Gaussian distribution is used for data more skewed than that suggested by the gamma distribution.

The mean-variance relationship: $V (\mu) = \mu^2$

Dispersion: The dispersion $??$ is unknown and needs to be estimated. MLE estimator cannot be found in closed form. The deviance is sensitive to very small values of $y_i$ for gamma EDMs, the Pearson estimator(R defualt) is recommended.

Link function: 1. inverse link function (canonical and default); 2. log link function; 3. identity link function. In practice, the log link function is more often used to ensure $u>0$ and for interpretaion purposes. Careful choice of the link function and transformations of the covariates can be used to describe asymptotic relationships between $y$ and $x$.

Inference: Wald, score or likelihood ratio test can be used.

The saddlepoint aprroximation is adequate for Gamma distribution when ?? ??? 1/3.

Overdispersion: 
1. the $m_i$ cases, of which observation $y_i$ is a proportion, are not independent. (Hierachical models)
2. the Bernoulli cases, that $m_i$ make up observation, are positively correlated. (Quasi-binomial models)

Diagnostic analysis of the model
standardized residuals against the fitted values
Q-Q plot
Cook's Distance plot
Influential observation


#### Data
The data set is available as the *blocks* data in the GLMsData library.

Children were seated at a small table and told to build a tower from the blocks
as high as they could. This was demonstrated for the child. The time taken and 
the number of blocks used were recorded. The cubes were always presented first, 
then cylinders. The second trial was conducted one month later.

The blocks were half inch cubes and cylinders included in Mrs. Hailmann's Beads
No. 470 of Bradley's Kindergarten Material. Throughout the article, the children
are referred to using male pronouns, but (in keeping with the custom at the
time) it is unclear whether all children were males or not. However, since 
gender is not recorded the children may all have been boys.

The source (Johnson and Courtney, 1931) gives the age in years and months. Here
they have been converted to decimal years.

#### Variables

Child: a child identifier from A to Y

Number: the number of blocks the child could successfully stack

Time: the time (in seconds) taken for the child to make the stack of blocks

Trial: the trial number on which the data were gathered

Shape: the shape of the blocks being stacked, either cubes or cylinders

Age: the age of the child (in years)

#### Research questions

*Note: We need our observations to be independent and there were two trials*
*in the study, so only use the first trial when analyzing these data.*

Is the shape of the block and age of the child related to the number of blocks
that they can put into a tower? Are the effects of age different for the 
different shapes?

Children who is one week older in age tend to use 2 more blcoks to build the tower. As for 
children of the same age, they use 1 more block to build the cubical tower than the 
cylindrical tower.



#### Analysis

1.Prepare the data.
```{r}
data(blocks)
str(blocks)
```

Subset the data and only select observations of participants' first trial
```{r}
block_ft <- subset(blocks, blocks$Trial == 1)
head(block_ft)
summary(block_ft)
```
Here is a histogram of the mass of the foliage (in kg). 

```{r}

hist(block_ft$Number,
     xlab = "Number of Blocks",
     main = "A histogram of the number of blocks")

```

2. Explore the relationships of each explanatory variable with the response variable. 

Let's start with Shape.

```{r}

plot(Number ~ Shape,
     data = block_ft,
     las = 1,
     xlab = "Shape",
     ylab = "Number of blocks")
```
Buiding cubical tower on average requires more blocks than cylindrical tower.


Now look at the relationship of number of blocks and age.
```{r}
plot(Number ~ Age,
     data = block_ft,
     las = 1,
     xlab = "Age (in years)",
     ylab = "Number of blocks")
```
The number seems to increase as age goes up.

3. Fit the model.


```{r}
block.log <- glm(Number ~ Shape * Age,
                family = Gamma(link = "identity"),
                data = block_ft)
anova(block.log, test = "F")
```


```{r}
block.log1 <- glm(Number ~ Age * Shape,
                family = Gamma(link = "identity"),
                data = block_ft)
anova(block.log1, test = "F")
```
We come to the same conclusion.


```{r}
block.log2 <- glm(Number ~ Shape * Age,
                family = Gamma(link = "identity"),
                data = block_ft)
```

Coefficients and effect sizes
```{r}
printCoefmat(coef(summary(block.log2)))
```
Children who is one week older in age tend to use 2 more blcoks to build the tower. As for children of the same age, they use 1 more block to build the cubical tower than the cylindrical tower.


4.Diagnostic Analysis

Quantile residual vs fitted values.

```{r}
qres <- qresid(block.log2)
scatter.smooth(qres ~ fitted(block.log2),
               las = 1,
               main = "Residuals vs fitted",
               xlab = "Fitted value",
               ylab = "Quantile residual")


```


Q-Q Plot
```{r}
qqnorm(rstandard(block.log2), las=1 )
qqline(rstandard(block.log2))
```

Cook's distance
```{r}
plot( cooks.distance(block.log2), type="h", las=1)
```
The plots are not bad.


Influential observations
```{r}
im <- influence.measures(block.log2)
colSums(im$is.inf)

```
No influential observations were detected according to Cook's distance test.


#### Additional exercise

Group ages into three categories and then use the six groupings (three ages by two blocks shapes) to study the mean-variance relationship. Show that the gamma distribution was a good choice for the random component of the model. (To help with your R code, there is a good example of this process on pages 429 and 430 in the textbook.)

```{r}
block_ft$AgeGroup <- cut(block_ft$Age, breaks = 3) 
vr <- with(block_ft, tapply(Number, list(AgeGroup, Shape), "var"))
mn <- with(block_ft, tapply(Number, list(AgeGroup, Shape), "mean"))

plot( log(vr) ~ log(mn), 
      las=1, 
      pch=19,
      xlab="log(group means)", 
      ylab="log(group variance)")

mf.lm <- lm( c(log(vr)) ~ c(log(mn)) )
coef( mf.lm )
abline( coef( mf.lm ), lwd=2)
```
The slope for the means is 2.27. The group variance is approximately proportional to square of the group mean. In other words, $V (\mu)$ ??? $\mu^2$ which corresponds to a gamma distribution.
