---
title: "EDRM878 Assignment3"
author: Tiejun Zhang
output: html_notebook
---
### 5.25. Children were asked to build towers as high as they could out of cubical and cylindrical blocks [2, 9]. The number of blocks used and the time taken were recorded (Table 2.12; data set: blocks). In this problem, only consider the number of blocks used y and the age of the child x.
```{r}
data(blocks)
summary(blocks)
```

#### 1.Plot the number of blocks used against the age of the child.
```{r}
scatter.smooth(jitter(blocks$Number) ~ blocks$Age, main = "Number of Blocks vs Age", xlab = "Age", ylab = "Number of blocks")
```


#### 2.From the plot and an understanding of the data, answer the two questions in Sect. 5.2 (p. 211) for these data, and hence propose a glm for the data.

What probability distribution is appropriate?

The plot indicates a non-linear relationship. Additionally, the variance tends to increases as the mean increases. As the response is the number of blocks(counts), the poisson distribution might be an appropriate option.


How are the explanatory variables related to the mean of the response $??$?
```{r}
# cut age into three groups
age.cut <- cut(blocks$Age, 3)
summary(age.cut)

# calculate the mean, varance and length for each age group.
mn <- tapply(blocks$Number, age.cut, "mean")
vr <- tapply(blocks$Number, age.cut, "var")
sz <- tapply(blocks$Number, age.cut, "length")

cbind("Group size" = sz, "Group mean" = mn, "Group variance" = vr)

```
As we can see from the output, the variance increases as the mean gets larger.

Side-by-side boxplots to present the trend.
```{r}
plot(jitter(Number) ~ (age.cut),
     pch = 1,
     las = 1,
     data = blocks,
     xlab = "Age Groups",
     ylab = "Number of blocks")


```

Plot of log of group means against log of group Variance
```{r}
plot(log(vr) ~ log(mn),
     pch = 19,
     las = 1,
     cex = 0.45 * sqrt(sz),
     xlab = "Log of means",
     ylab = "Log of variances")
```

The variance as a function of the mean.
```{r}
hm.lm <- lm(log(vr) ~ log(mn), weights = sz)
coef(hm.lm)
confint(hm.lm)

# dispersion index
exp(coef(hm.lm)[1])
```
The dispersion index is 0.08 and the variance function is $mean$^2.1.Given the data is discrete in nature, poisson distribution is appropriate.


### 6.10. Children were asked to build towers as high as they could out of cubical and cylindrical blocks [3, 6]. The number of blocks used and the time taken were recorded (data set: blocks). In this problem, only consider the number of blocks used y and the age of the child x. In Problem 5.25, a glm was proposed for these data.
#### 1.Fit this glm using r, and write down the fitted model.
```{r}
m1 <- glm(Number~Age, data=blocks, family=poisson(link = "log"))
m1
```

#### 2.Determine the standard error for each regression parameter.
```{r}
summary(m1)
```
Both coefficients (beta0 and beta1) have a p value less than 0.05, indicating a strong relationship. The standard error is 0.22 for beta0 and 0.05 for beta1.

#### 3.Compute the residual deviance.

```{r}
deviance(m1)
```

The residual deviance is 62.24.


***

**Nicely done!**

**20/20**
