---
title: "EDRM878 Homework 2"
author: Tiejun Zhang
output: html_notebook
---


### 3.16.Children were asked to build towers out of cubical and cylindrical blocks as high as they could [20, 33], and the number of blocks used and the time taken were recorded. The data (data set: blocks) and a fuller description are given in Problem 2.18 (p. 88). In that problem, a linear regression model was fitted to model the time to build the towers, based on the initial examination in Problem 1.9 (p. 28).

```{r}
library(GLMsData)
data(blocks)
str(blocks)
summary(blocks)
```


#### 1.Perform a diagnostic analysis of the linear regression model fitted in Problem 2.18 (p. 88), and show a transformation of the response is necessary.

```{r}
# fit model
linear.model <-lm(Time~Shape, data = blocks)
summary(linear.model)
```
The model only has one explanatory variable "Shape". It is a factor of two levels: Cube and Cylinder.

```{r}
# linearity check
plot(rstandard(linear.model) ~ blocks$Shape, ylab = "Standardized residuals", xlab = "Shape", main = "Fig.1 Standardized Residuals against Shape")
```
Fig1 suggests that the relationship between the residuals and Shape is not linear, for Shape is a two-level factor.

```{r}
# Constant variance check
plot(rstandard(linear.model) ~ fitted(linear.model), ylab = "Standardized residuals", xlab = "Fitted values", main = "Fig.2 Standardized Residuals against Fitted Values")
```
Fig2 shows that the variance differs between cubes and cylinders. The cube group is less variant than the cylinder group.

```{r}
# normality check
qqnorm(rstandard(linear.model), las = 1, pch = 19, main = "Fig.3 Standardized Residuals against Normal Quantiles")
qqline(rstandard(linear.model))
```

Fig3 indicates the residuals are not normally distributed. The distribution has a heavy up tail. 

```{r}
# Influential Observations
inf_ms <- influence.measures(linear.model)

# Influential on three criteria
rowSums(inf_ms$is.inf)
table(rowSums(inf_ms$is.inf))

# Influential observations
inf_ms$is.inf[c(6, 76),]


# Plots
# Cook,s distance
plot(cooks.distance(linear.model), type = "h", main = "Cook's distance", ylab = "D", xlab = "Observation number", las = 1)

# DFFITS
plot(dffits(linear.model), type = "h", main = "DFFITS", ylab = "DFFITS", xlab = "Observation number", las = 1)

# DFBETAS
plot(dfbetas(linear.model)[, 2], type = "h", main = "DFBETAS for beta2", ylab = "DFBETAS", xlab = "Observation number", las = 1)

```
98 observations are not declared influential on any of the criteria; Observation 6 is identified influnetial on all three criteria and observation 76 is declared influential on two criteria.

The residuals show non-normality, non-constant variance and two influential observations, therefore, it is necessary to transform the response variable.


#### 2.Fit an appropriate linear regression model to the data after applying the transformation, ensuring a diagnostic analysis.

```{r}
# fit model
m0 <- lm(log(Time) ~ 1, data = blocks)
m1 <- update(m0, . ~ . + Shape*Age)
anova(m1)

t.model <- lm(log(Time) ~ Shape, data = blocks)

```
The sequential tests suggest that only Shape is an effective predictor (p < 0.05) for the log-tranformed response.

Diagnostic Analysis
```{r}
plot(rstandard(t.model) ~ blocks$Shape, ylab = "Standardized residuals", xlab = "Shape", main = "Fig.4 Standardized Residuals against Shape")
```
Fig4 still shows a non-linear relationship but with less extreme points on up tail.

```{r}
# Constant variance check
plot(rstandard(t.model) ~ fitted(linear.model), ylab = "Standardized residuals", xlab = "Fitted values", main = "Fig.5 Standardized Residuals against Fitted Values")
```
Fig5 shows the variance appear to be constant between two shapes.


```{r}
# normality check
qqnorm(rstandard(t.model), las = 1, pch = 19, main = "Fig.6 Standardized Residuals against Normal Quantiles")
qqline(rstandard(t.model))
```
Fig6 indicates that the residuals are quite close to a normal distribution.

```{r}
# Influential Observations
trf_inf_ms <- influence.measures(t.model)

# Influential on three criteria
rowSums(trf_inf_ms$is.inf)
table(rowSums(trf_inf_ms$is.inf))

# Influential observations
trf_inf_ms$is.inf[c(6, 76),]

# Plots
# Cook's distance
plot(cooks.distance(t.model), type = "h", main = "Cook's distance", ylab = "D", xlab = "Observation number", las = 1)

# DFFITS
plot(dffits(t.model), type = "h", main = "DFFITS", ylab = "DFFITS", xlab = "Observation number", las = 1)

# DFBETAS
plot(dfbetas(t.model)[, 2], type = "h", main = "DFBETAS for beta2", ylab = "DFBETAS", xlab = "Observation number", las = 1)

```
98 observations are not declared influential on any of the criteria; Observation 6 is identified influnetial on two criteria and observation 76 is declared influential on just one criterion.

Overall, the model fit improved after a log-tranform of the response.

***

**Very nicely done!**

**20/20**

